Designing a high-performance boundary element library with OpenCL and Numba

### 2021-04-06 bem.tex
start working through opencl section

\section{Boundary element methods with Bempp}
In this section, we provide a brief introduction to boundary element methods and describe the necessary steps for their numerical discretization and solution.

>>>>

\section{BOUNDARY ELEMENT METHODS WITH BEMPP}
In this section, we provide a brief introduction to boundary element methods and describe the necessary steps for their numerical discretization and solution.

### 2021-04-06 main.tex
Finished most of the OpenCL section.
edititing first section
Make tikz version of figure
add command for \ds
working on bempp section
american spelling
start working through opencl section

\usepackage{upmath}
\usepackage{amsmath, amssymb}

\def\bx{\mathbf{x}}
\def\by{\mathbf{y}}

\jvol{XX}
\jnum{XX}
\paper{8}
\jmonth{May/June}
\jname{IT Professional}
\pubyear{2019}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

@@

\begin{abstract}
The Bempp boundary element library is a well known library for the simulation of a range of electrostatic, acoustic and electromagnetic problems in homogeneous bounded and unbounded domains. It originally started as a traditional C++ library with a Python interface. Over the last two years we have completely redesigned Bempp as a native Python library, called Bempp-cl, that provides computational backends for OpenCL (using PyOpenCL) and Numba. The OpenCL backend implements kernels for GPUs and SIMD optimised for CPUs. In this paper we will discuss the design of Bempp-cl, provide performance comparisons on different compute devices and discuss the advantages and disadvantages of OpenCL as compared to Numba.
\end{abstract}

@@

\input{intro}
\input{bem}

>>>>

\usepackage{xcolor}
\definecolor{bemppgrey}{HTML}{555555}
\definecolor{bempporange}{HTML}{FF7A1A}

\usepackage{pythonhighlight}

\usepackage{tikz}
\usetikzlibrary{arrows,decorations.pathreplacing}

\usepackage{ifthen}

\usepackage{upmath}
\usepackage{upgreek}
\usepackage{amsmath, amssymb}

\usepackage{cleveref}
\crefname{equation}{Equation}{Equations}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}
\crefname{section}{Section}{Sections}
\Crefname{section}{Section}{Sections}

% Bold vector (for points in R^3)
\newcommand{\bvec}[1]{\boldsymbol{#1}}
\def\bx{\bvec{x}}
\def\by{\bvec{y}}
\newcommand{\ds}[1][]{\,\mathrm{d}s\ifthenelse{\equal{#1}{}}{}{_{#1}}}

% Discrete vector and discrete matrix
\newcommand{\dmat}[1]{\mathbf{#1}}
\newcommand{\dvec}[1]{\mathbf{#1}}

\jvol{XX}
\jnum{XX}
\paper{8}
\jmonth{May/June}
\jname{IT Professional}
\pubyear{2021}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

@@

\begin{abstract}
The Bempp boundary element library is a well known library for the simulation of a range of electrostatic, acoustic and electromagnetic problems in homogeneous bounded and unbounded domains. It originally started as a traditional C++ library with a Python interface. Over the last two years we have completely redesigned Bempp as a native Python library, called Bempp-cl, that provides computational backends for OpenCL (using PyOpenCL) and Numba. The OpenCL backend implements kernels for GPUs and SIMD optimized for CPUs. In this paper we will discuss the design of Bempp-cl, provide performance comparisons on different compute devices and discuss the advantages and disadvantages of OpenCL as compared to Numba.
\end{abstract}

@@

\input{intro}
\input{bem}
\input{bempp_overview}
\input{opencl}

\bibliographystyle{IEEEtran}
\begin{thebibliography}{1}

\bibitem{Bempp-cl}
T. Betcke and M. W. Scroggs, ``Bempp-cl: A fast Python based just-in-time compiling boundary element library,''
{\it Journal of Open Source Software}, vol. 6, no. 59, p. 2879, 2021.
\textsc{doi}: \href{https://dx.doi.org/10.21105/joss.02879}{10.21105/joss.02879}
\end{thebibliography}

### 2021-04-06 opencl.tex
OpenCL highlighting
american spelling
start working through opencl section

\section{Assembling boundary integral operators with OpenCL}

In this section we discuss in more detail the assembly of boundary integral operators with OpenCL
and how we integrated this into our Python workflow. We start with a brief introduction to OpenCL and then
dive into how we use OpenCL as part of Bempp-cl.

@@

OpenCL (\url{https://www.khronos.org/opencl/}) is a heterogeneous compute standard for CPUs, GPUs, FPGAs, and other types of devices that provide conformant drivers. At its core OpenCL executes compute kernels that can be written in C99, or more recently, in C++. The current version of OpenCL is 3.0, though the most widely implemented standard is OpenCL 1.2, which Bempp-cl uses. OpenCL splits the computational tasks into work-items, each of which represents a single execution of a compute kernel. Work items are grouped together into work-groups, which share local memory. Barrier synchronization is only allowed within a work-group. All work-items are uniquely indexed by a one, two, or three dimensional index space, called NDRange. Kernels are launched onto a compute device (e.g. CPU or GPU) from a host device. OpenCL allows kernels to be loaded as strings and compiled on the fly for a given device, making it well suited for launching from high-productivity languages. To launch an OpenCL kernel the following steps need to be performed.
\begin{itemize}
	\item Initialize buffers on the compute device.
	\item Compute relevant data from host to the compute device.
	\item Load a kernel string and just-in-time compile it for the device.
	\item Execute the kernel for the given buffers and chosen index space.
	\item Copy results back to the host.
\end{itemize}
An excellent feature of OpenCL is its very good support for vectorized operations. OpenCL provides vector data types, such as \textbf{double4}, that can hold four double values in a SIMD register and defines a number of standard operations for these vector types. This makes it easy to explicitly target modern SIMD execution in a portable way while avoiding difficult compiler intrinsics and keeping kernel code readable.

@@

\begin{figure*}
	\center
	\includegraphics[width=12cm]{img/kernel_header}
	\caption{Definition of the OpenCL compute kernel for scalar integral equations.}
	\label{fig:kernel_definition}
\end{figure*}

@@

\begin{itemize}
	\item Test and trial indices of triangles over which to be integrated.
	\item Signs of the normal directions for the spaces.
	\item Test and trial grid as flat floating point array, defining each triangle through nine floating point numbers, specifying the $(x, y, z)$ coordinates of each of the three nodes of a triangle.
	\item Test and trial connectivity information that stores the indices of each node for each triangle.
	\item Test and trial mappings of local triangle degrees of freedom to global degrees of freedom.
	\item Test and trial basis function multipliers for each triangle.
	\item Quadrature points and quadrature weights.
	\item A buffer that contains the global assembled matrix.
	\item Additional kernel parameters, such as the wavenumber for Helmholtz problems.
	\item Number of test and trial degrees of freedom.
	\item A single byte that is set to one if the test and trial grids are disjoint.
\end{itemize}
The actual kernel definition is shown in Figure \ref{fig:kernel_definition}.
Before the kernel can be launched, it needs to be configured and just-in-time compiled. Kernel configuration happens through C-style preprocessor definitions that are passed through the just-in-time compiler. These include the names of the test and trial space, the name of the function that evaluates the integral kernel, choosing between single and double precision types, and for SIMD enhanced kernel the vector length of the SIMD types.
For example, in Figure \ref{fig:kernel_definition} all floating point types have the name REALTYPE. This is substituted with either \textit{float} or \textit{double} during just-in-time compilation.

@@

\section{Numba assembly of integral operators}

@@

\section{SIMD optimized evaluation of the Laplace kernel with Numba}

>>>>

\section{ASSEMBLING BOUNDARY INTEGRAL OPERATORS WITH OPENCL}

In this section, we discuss in more detail the assembly of boundary integral operators with OpenCL
and how we integrated this into our Python workflow. We start with a brief introduction to OpenCL and then
dive into how we use OpenCL as part of Bempp-cl.

@@

OpenCL (\url{https://www.khronos.org/opencl/}) is a heterogeneous compute standard for CPUs, GPUs, FPGAs, and other types of devices that provide conformant drivers. At its core, OpenCL executes compute kernels that can be written in C99, or (more recently) in C++. The current version of OpenCL is 3.0, though the most widely implemented standard is OpenCL 1.2, which Bempp-cl uses.

OpenCL splits computational tasks into work-items, each of which represents a single execution of a compute kernel. Work items are grouped together into work groups, which share local memory. Barrier synchronization is only allowed within a work group. All work items are uniquely indexed by a one, two, or three dimensional index space, called NDRange. Kernels are launched onto a compute device (e.g., a CPU or GPU) from a host device. OpenCL allows kernels to be loaded as strings and compiled on the fly for a given device, making it well suited for launching from high-productivity languages.

To launch an OpenCL kernel the following steps need to be performed:
\begin{itemize}
	\item Initialize buffers on the compute device.
	\item Compute relevant data from host to the compute device.
	\item Load a kernel string and just-in-time compile it for the device.
	\item Execute the kernel for the given buffers and chosen index space.
	\item Copy results back to the host.
\end{itemize}
An excellent feature of OpenCL is its very good support for vectorized operations. OpenCL provides vector data types, such as \textbf{double4}, that can hold four double values in a SIMD register and defines a number of standard operations for these vector types. This makes it easy to explicitly target modern SIMD execution in a portable way while avoiding difficult compiler intrinsics and keeping kernel code readable.

@@

\begin{figure*}
	\center
	\input{img/kernel_header}
	\caption{Definition of the OpenCL compute kernel for scalar integral equations.}
	\label{fig:kernel_definition}
\end{figure*}

@@

\begin{itemize}
	\item Test and trial indices of triangles over which to be integrated.
	\item Signs of the normal directions for the spaces.
	\item Test and trial grid as flat floating point array, defining each triangle through nine floating point numbers, specifying the $(x, y, z)$ coordinates of each of the three nodes of a triangle.
	\item Test and trial connectivity information that stores the indices of each node for each triangle.
	\item Test and trial mappings of local triangle degrees of freedom to global degrees of freedom.
	\item Test and trial basis function multipliers for each triangle.
	\item Quadrature points and quadrature weights.
	\item A buffer that contains the global assembled matrix.
	\item Additional kernel parameters, such as the wavenumber for Helmholtz problems.
	\item Number of test and trial degrees of freedom.
	\item A single byte that is set to one if the test and trial grids are disjoint.
\end{itemize}
The actual kernel definition is shown in Figure \ref{fig:kernel_definition}.
Before the kernel can be launched, it needs to be configured and just-in-time compiled. Kernel configuration happens through C-style preprocessor definitions that are passed through the just-in-time compiler. These include the names of the test and trial space, the name of the function that evaluates the integral kernel, choosing between single and double precision types, and for SIMD enhanced kernel the vector length of the SIMD types.
For example, in Figure \ref{fig:kernel_definition} all floating point types have the name \ocl{REALTYPE}. This is substituted with either \textit{float} or \textit{double} during just-in-time compilation.

@@

\section{NUMBA ASSEMBLY OF INTEGRAL OPERATORS}

@@

\section{SIMD OPTIMIZED EVALUATION OF THE LAPLACE KERNEL WITH NUMBA}
